---
title: "Module 6.1"
subtitle: "Logistic Regression"
format:
  html:
    code-link: true
highlight-style: atom-one
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-tip}
## Prework

Install the `peacesciencer` package and have a look at its [documentation](https://svmiller.com/peacesciencer/) to familiarize yourself with its contents and basic functions:

```r
install.packages("peacesciencer")
```
:::

## Overview

So far, we have modeled continuous outcomes like democracy scores and GDP per capita. But many of the most interesting questions in research involve **binary outcomes** — yes or no, conflict or no conflict, voted or didn't vote. Logistic regression is the standard tool for modeling these binary outcomes, and it is one of the most widely used methods in the social sciences.

{{< video https://youtu.be/p-ltr1C7u2o?si=TLb3y75pSaJ067ue title='Classification: Motivation' >}}

## Binary Outcomes

A **binary outcome** (or binary response variable) is a variable that takes on exactly two possible values, typically coded as 0 and 1. Examples include:

- Did civil conflict begin in this country this year? (0 = No, 1 = Yes)
- Did a country vote for this UN resolution? (0 = No, 1 = Yes)
- Did a voter turn out in the election? (0 = No, 1 = Yes)
- Does a country have judicial review? (0 = No, 1 = Yes)

For each observation, we are interested in modeling the **probability** that the outcome equals 1. This is fundamentally different from predicting a continuous outcome.

## Why Not Linear Regression?

You might wonder: why not just use the linear regression we already know? The problem is that linear regression can produce **impossible predictions**. Probabilities must be between 0 and 1, but a linear model can predict values like −0.3 or 1.7 — which are mathematically meaningless as probabilities.

Logistic regression solves this by using a mathematical transformation that ensures all predicted values remain between 0 and 1, no matter what the predictor values are.

## Bernoulli Trials and Probability Modeling

The right framework for binary outcomes is the **Bernoulli trial**: a random experiment with exactly two possible outcomes (success and failure), where success is coded as 1. When we model civil conflict onset, each country-year is a separate Bernoulli trial:

$$y_i \sim \text{Bernoulli}(p_i)$$

Each observation has its own probability $p_i$ of "success" (conflict onset in this case — the term is statistical, not normative). Our goal is to model how predictor variables influence this probability.

## Generalized Linear Models

Logistic regression belongs to a broader family of models called **Generalized Linear Models (GLMs)**. All GLMs share three characteristics:

1. **A probability distribution** describing the outcome (Bernoulli for binary data)
2. **A linear predictor**: $\eta_i = \beta_0 + \beta_1 X_{1i} + \cdots + \beta_k X_{ki}$
3. **A link function** that connects the linear predictor to the outcome probability

The link function is the key innovation: it translates between the unconstrained linear predictor (which can range from −∞ to +∞) and the probability (which must stay between 0 and 1).

## The Logistic Regression Model

{{< video https://youtu.be/xuTiAW0OR40?si=sUoIr5pzHwHEpazO title='Logistic Regression' >}}

### The Sigmoid Function

The **sigmoid function** (also called the logistic function) takes any real number and converts it to a probability between 0 and 1:

$$\sigma(z) = \frac{1}{1 + e^{-z}}$$

```{r}
#| label: sigmoid-function
#| echo: false

library(tidyverse)

sigmoid_data <- tibble(z = seq(-6, 6, length.out = 1000)) |>
  mutate(sigma_z = 1 / (1 + exp(-z)))

ggplot(sigmoid_data, aes(x = z, y = sigma_z)) +
  geom_line(linewidth = 1, color = "steelblue") +
  ylim(0, 1) +
  xlab("z (linear predictor)") +
  ylab("σ(z) = probability") +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "The Sigmoid Function",
       subtitle = "Converts any real number to a probability between 0 and 1") +
  theme_minimal()
```

The characteristic S-shape guarantees the output stays between 0 and 1. When the linear predictor is 0, the probability is exactly 0.5.

### The Logit Link Function

The logistic regression model uses the **logit** as its link function. For $0 \leq p \leq 1$:

$$\text{logit}(p) = \log\left(\frac{p}{1-p}\right)$$

The logit is the log of the **odds** ($p$ divided by $1-p$). It maps probabilities from [0, 1] to the entire real line, which is exactly what we need to model them with a linear predictor:

```{r}
#| label: logit-function
#| echo: false

logit_data <- tibble(p = seq(0.001, 0.999, length.out = 1000)) |>
  mutate(logit_p = log(p / (1 - p)))

ggplot(logit_data, aes(x = p, y = logit_p)) +
  geom_line(linewidth = 1, color = "coral") +
  xlim(0, 1) +
  ylab("logit(p)") +
  xlab("p (probability)") +
  labs(title = "The Logit Function",
       subtitle = "Converts probabilities to an unconstrained scale") +
  theme_minimal()
```

Putting it all together, logistic regression models:

$$\text{logit}(p_i) = \beta_0 + \beta_1 X_{1i} + \cdots + \beta_k X_{ki}$$

To get predicted probabilities, we apply the sigmoid function to the linear predictor.

## Analyzing Conflict Onset

We will use the `peacesciencer` package to examine what factors are associated with civil conflict onset. The package provides convenient functions to build a dataset of country-years and add various predictors. We replicate part of Fearon and Laitin's (2003) influential study of civil war.

```{r}
#| label: conflict-data

library(peacesciencer)

conflict_df <- create_stateyears(system = 'gw') |>
  filter(year %in% c(1946:1999)) |>
  add_ucdp_acd(type = c("intrastate"), only_wars = FALSE) |>
  add_democracy() |>
  add_creg_fractionalization() |>
  add_sdp_gdp() |>
  add_rugged_terrain()

glimpse(conflict_df)
```

Each row represents one country in one year. The variable `ucdponset` is our binary outcome — it equals 1 if a civil conflict began in that country in that year, and 0 otherwise.

### Running a Logistic Regression

The syntax for logistic regression is nearly identical to linear regression — we switch from `lm()` to `glm()` and add `family = "binomial"` to specify the logit link:

```{r}
#| label: bivariate-model

conflict_model <- glm(ucdponset ~ wbgdppc2011est,
                      data = conflict_df,
                      family = "binomial")

summary(conflict_model)
```

### Interpreting the Coefficients

The model equation on the log-odds scale is:

$$\log\left(\frac{p}{1-p}\right) = -1.16 - 0.33 \times \text{log GDP per capita}$$

The coefficient for log GDP per capita is negative and significant — wealthier countries have lower odds of conflict onset. But interpreting the magnitude on the log-odds scale is not intuitive, so we convert to **odds ratios** by exponentiating the coefficient.

### Odds Ratios

{{< video https://youtu.be/oIR8YOSXudM?si=aEALe2ifW71lsJJG title='Understanding Odds Ratios' >}}

```{r}
#| label: odds-ratios

exp(coef(conflict_model))
```

The odds ratio for log GDP per capita is approximately 0.718. This means that for each one-unit increase in log GDP per capita, the **odds** of conflict onset are multiplied by 0.718 — a decrease of about 28%. An odds ratio less than 1 indicates a negative association; greater than 1 indicates a positive association.

## Predicted Probabilities

Odds ratios tell us about relative effects, but for concrete interpretation we want **predicted probabilities**: what is the probability of conflict onset for a country with a given level of GDP?

We can use the `marginaleffects` package to calculate these for specific cases. Let's predict conflict onset for the United States, Venezuela, and Rwanda in 1999:

```{r}
#| label: marginal-effects

library(marginaleffects)

selected_countries <- conflict_df |>
  filter(
    gw_name %in% c("United States of America", "Venezuela", "Rwanda"),
    year == 1999
  )

marg_effects <- predictions(conflict_model, newdata = selected_countries)

tidy(marg_effects) |>
  select(estimate, p.value, conf.low, conf.high, gw_name)
```

The predicted probability of conflict onset is highest for Rwanda, followed by Venezuela, and lowest for the United States — consistent with the negative relationship between wealth and conflict onset.

::: {.callout-warning icon=false}
## Your Turn

Using the `conflict_df` dataset, try running a logistic regression with `ucdponset` as the outcome and `v2x_polyarchy` (the V-Dem democracy measure) as the predictor.

- Is the coefficient positive or negative? What does this suggest about the relationship between democracy and conflict onset?
- Exponentiate the coefficient to get the odds ratio and interpret it.
- Use `predictions()` from `marginaleffects` to calculate the predicted probability of conflict onset for three countries with different democracy levels.
:::
