---
title: "Module 2.1"
subtitle: "Working with and Summarizing Data"
format:
  html:
    code-link: true
highlight-style: atom-one
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-tip}
## Prework

Generate a Quarto document named `module-2.1.qmd` in your modules project folder so that you can code along with the lesson.
:::

## Overview

In this module we learn how to download, wrangle, and summarize data. We will work with the [V-Dem Dataset](https://www.v-dem.net/data/the-v-dem-dataset/), a rich source of democracy-related indicators covering countries from 1789 to the present. Along the way we cover the core `dplyr` verbs for data manipulation: `filter()`, `select()`, `mutate()`, `group_by()`, `summarize()`, and `arrange()`.

## Downloading and Transforming Data

{{< video https://youtu.be/ScQD47eJeAE title='Downloading Data from V-Dem' >}}

The `vdemdata` package gives us direct access to the V-Dem dataset in R. Its main object, `vdem`, contains the entire dataset — over 4,000 variables and nearly 30,000 country-year observations. Because it downloads everything at once, we rely on `dplyr` functions to narrow it down to just what we need.

::: {.callout-note}
V-Dem has a variable look-up tool (`find_var`), but for details on each indicator it is best to consult the [V-Dem codebook](https://www.v-dem.net/documents/24/codebook_v13.pdf) directly.
:::

We will focus on four variables: the polyarchy score (`v2x_polyarchy`), GDP per capita (`e_gdppc`), region (`e_regionpol_6C`), and the women's empowerment index (`v2x_gender`). We also keep `country_name`, `country_id`, and `year` for identification.

The `filter()` verb keeps only the rows we want. The `select()` verb keeps only the columns we want — and lets us rename them at the same time. Finally, `mutate()` with `case_match()` recodes the numeric region codes into readable labels.

```{r}
#| label: democracy

library(tidyverse)
library(vdemdata)

democracy <- vdem |>
  filter(year >= 1990) |>
  select(
    country = country_name,
    vdem_ctry_id = country_id,
    year,
    polyarchy = v2x_polyarchy,
    gdp_pc = e_gdppc,
    women_emp = v2x_gender,
    region = e_regionpol_6C
  ) |>
  mutate(
    region = case_match(region,
      1 ~ "Eastern Europe",
      2 ~ "Latin America",
      3 ~ "Middle East",
      4 ~ "Africa",
      5 ~ "The West",
      6 ~ "Asia")
  )

glimpse(democracy)
```

::: {.callout-warning icon=false}
## Your Turn

- Change the start year from 1990 to a year of your choosing.
- Add one more V-Dem variable to the `select()` call. Look up a variable code in the codebook that interests you and give it a clean name.
- Run `glimpse()` on your new data frame. How many rows and columns does it have?
:::

## Summarizing Data

{{< video https://youtu.be/mgBwVzMV-v8 title='Group, Summarize and Arrange' >}}

A common workflow in data science is **group → summarize → arrange**. We group the data by a categorical variable, calculate summary statistics for each group, and then sort the results.

We will illustrate this with `dem_women.csv`, a pre-processed dataset combining V-Dem democracy indicators with economic and gender representation variables.

```{r}
#| label: dem-women

dem_women <- read_csv("data/dem_women.csv")

glimpse(dem_women)
```

Now let's calculate regional averages for all four variables and sort by polyarchy score in descending order:

```{r}
#| label: dem-summary

dem_summary <- dem_women |>
  group_by(region) |>
  summarize(
    polyarchy   = mean(polyarchy, na.rm = TRUE),
    gdp_pc      = mean(gdp_pc, na.rm = TRUE),
    flfp        = mean(flfp, na.rm = TRUE),
    women_rep   = mean(women_rep, na.rm = TRUE)
  ) |>
  arrange(desc(polyarchy))

dem_summary
```

`group_by()` splits the data into groups before the summarize step. `summarize()` collapses each group down to one row. `arrange()` sorts the result — `desc()` reverses the default ascending order.

We can export the summary for later use:

```{r}
#| label: export
#| eval: false

write_csv(dem_summary, "data/dem_summary.csv")
```

::: {.callout-warning icon=false}
## Your Turn

- Summarize the `dem_women` data by region using `median()` instead of `mean()`. Do the rankings change?
- Add a new summary statistic: the standard deviation of `polyarchy` using `sd()`. Which region has the most variation in democracy scores?
- Try `arrange(polyarchy)` (without `desc()`). What changes?
:::
