---
title: "Module 5.2"
subtitle: "Multiple Regression"
format:
  html:
    code-link: true
highlight-style: atom-one
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-tip}
## Prework

Run this code chunk to load the necessary packages and data for this module:

```{r}
#| label: setup
#| code-fold: true

library(tidyverse)
library(vdemdata)

model_data <- vdem |>
  filter(year == 2006) |>
  select(
    country = country_name,
    libdem = v2x_libdem,
    wealth = e_gdppc,
    oil_rents = e_total_oil_income_pc,
    polarization = v2cacamps,
    corruption = v2x_corr,
    judicial_review = v2jureview_ord,
    region = e_regionpol_6C,
    regime = v2x_regime
  ) |>
  mutate(
    log_wealth = log(wealth),
    judicial_review = factor(judicial_review, labels = c("No", "Yes")),
    region = factor(
      region,
      labels = c(
        "Eastern Europe",
        "Latin America",
        "MENA",
        "Sub-Saharan Africa",
        "Western Europe & North America",
        "Asia & Pacific"
      )
    )
  )

glimpse(model_data)
```
:::

## Overview

In the previous module, we discussed simple linear regression, which models the relationship between a response variable and a single explanatory variable. In this module, we extend that to **multiple linear regression**, which allows us to model the relationship between a response variable and *multiple* predictors simultaneously.

{{< video https://youtu.be/jXg0vU0y1ak?si=rhQ9uWhuhVqnI-Ox title='Multiple Linear Regression' >}}

## Multiple Regression

Multiple linear regression extends the bivariate model by adding more predictors:

$$Y = a + b_1X_1 + b_2X_2 + \ldots + b_nX_n$$

Where $Y$ is the response variable, $a$ is the intercept, and $b_1$ through $b_n$ are the coefficients for predictors $X_1$ through $X_n$.

The key phrase in multiple regression is **"holding other variables constant"** (sometimes phrased as *ceteris paribus*, Latin for "all else equal"). Each coefficient represents the expected change in $Y$ for a one-unit change in that predictor, *while controlling for all the other predictors in the model*.

We use data from 2006 to demonstrate. Let's start by looking at polarization as a predictor of democracy:

```{r}
#| label: plot-polarization

ggplot(model_data, aes(x = polarization, y = libdem)) +
  geom_point() +
  geom_smooth(method = "lm", color = "#E48957", se = FALSE) +
  labs(x = "Polarization", y = "Liberal Democracy Index") +
  theme_minimal()
```

There is a clear negative relationship — more polarized countries tend to be less democratic. Let's estimate this bivariate model:

```{r}
#| label: model1

model1 <- lm(libdem ~ polarization, data = model_data)

summary(model1)
```

The coefficient for polarization is approximately −0.088 and statistically significant: a one-unit increase in polarization is associated with a 0.088 decrease in the democracy score.

Now let's add log wealth as a second predictor:

$$lib\_dem = a + b_1 \times polarization + b_2 \times log(wealth)$$

```{r}
#| label: model2

model2 <- lm(libdem ~ polarization + log_wealth, data = model_data)

summary(model2)
```

The coefficient for polarization is now approximately −0.05. This is the effect of polarization **holding GDP per capita fixed**. Similarly, the coefficient for log wealth tells us the association with democracy **holding polarization fixed**. Notice how the coefficients can change when we add predictors — this happens when predictors are correlated with each other.

Let's add a third predictor — oil rents per capita:

$$lib\_dem = a + b_1 \times polarization + b_2 \times log(wealth) + b_3 \times oil\_rents$$

```{r}
#| label: model3

model3 <- lm(libdem ~ polarization + log_wealth + oil_rents, data = model_data)

summary(model3)
```

The oil rents coefficient is negative and significant: **holding polarization and GDP fixed**, higher oil rents are associated with lower democracy.

## Categorical Predictors

Another important topic is how to include **categorical variables** in regression models. Categorical variables represent distinct groups rather than continuous numeric values — examples include regime type, world region, or the presence or absence of an institution.

We handle categorical variables through **dummy variables** (also called indicator variables): one binary (0/1) variable for each category, with one category left out as the **reference category**. R handles this automatically when a variable is stored as a `factor`.

### Binary Categorical Variable: Judicial Review

Let's explore whether countries with judicial review tend to be more democratic. We already converted `judicial_review` to a factor in the setup chunk. Let's visualize the relationship:

```{r}
#| label: judicial-review-plot

ggplot(model_data, aes(x = wealth, y = libdem, color = judicial_review)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10(labels = scales::dollar_format(suffix = "k")) +
  labs(x = "GDP per capita", y = "Liberal Democracy Index") +
  scale_color_manual(name = "Judicial Review", values = c("steelblue3", "coral"),
                     labels = c("No", "Yes")) +
  theme_bw()
```

At every level of wealth, countries with judicial review tend to have higher democracy scores. We can quantify this by including `judicial_review` as a predictor:

```{r}
#| label: model4

model4 <- lm(libdem ~ log_wealth + judicial_review, data = model_data)

summary(model4)
```

::: {.callout-important}
## Understanding Reference Categories

When R includes a factor in a regression, it creates dummy variables for each level *except one*. That omitted level is the **reference category** (or baseline). All coefficients for the factor are then interpreted relative to that baseline.

Here, "No" is the reference category for `judicial_review` (R uses the first factor level). The intercept represents the expected democracy level for countries *without* judicial review. The coefficient for `judicial_reviewYes` tells us how much higher the expected democracy score is for countries *with* judicial review, holding GDP fixed.
:::

The coefficient of approximately 0.26 for `judicial_reviewYes` means that, **holding GDP per capita fixed**, countries with judicial review are expected to score about 0.26 points higher on the liberal democracy index.

### Multi-Category Variable: World Region

When a categorical variable has more than two levels, R creates one dummy for each level except the reference. Let's fit a model with world region as the predictor:

```{r}
#| label: region-levels

levels(model_data$region)
```

Eastern Europe is the first level, so it will serve as the reference category.

```{r}
#| label: model5

model5 <- lm(libdem ~ region, data = model_data)

summary(model5)
```

Each regional coefficient represents the expected difference in democracy between that region and Eastern Europe. For example, the coefficient for MENA tells us how much more (or less) democratic MENA countries are expected to be compared to Eastern European countries.

What if you want a different baseline? Use `relevel()` to change the reference category:

```{r}
#| label: model6

model_data_relevel <- model_data |>
  mutate(region2 = relevel(region, ref = "Sub-Saharan Africa"))

model6 <- lm(libdem ~ region2, data = model_data_relevel)

summary(model6)
```

Now all coefficients are interpreted relative to Sub-Saharan Africa. Eastern Europe is expected to be about 0.14 units more democratic than Sub-Saharan Africa, and so on.

## Making Predictions

Once we have estimated a model, we can use it to make predictions for hypothetical cases using R's `predict()` function. Let's use our model with polarization, wealth, and oil rents (model3). Suppose we want to predict the democracy level for a country with:

- Polarization = 0.5
- GDP per capita of ~$8k (log_wealth ≈ 2.1)
- Oil rents = 2

```{r}
#| label: predictions

new_data <- tibble(
  polarization = 0.5,
  log_wealth   = 2.1,
  oil_rents    = 2
)

predict(model3, newdata = new_data)
```

The model predicts a democracy score of approximately 0.46 for this hypothetical country. This prediction is a principled estimate based on the relationships learned from the original data — not a guarantee, but a useful summary of what the model expects.

::: {.callout-warning icon=false}
## Your Turn

The data also includes a categorical `regime` variable: Closed Autocracy (0), Electoral Autocracy (1), Electoral Democracy (2), Liberal Democracy (3), and a `corruption` measure.

- Convert `regime` into a factor in `model_data`.
- Check the levels and identify the reference category.
- Visualize the average corruption level across regime types using a bar plot.
- Fit a regression model predicting `corruption` from `regime`. Interpret the coefficients.
- Switch the reference category to "Electoral Democracy" (level 2) and refit. How do the coefficients change?
:::
