---
title: "Module 2.2"
subtitle: "Exploring Categorical Data"
format:
  html:
    code-link: true
highlight-style: atom-one
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-tip}
## Prework

Copy and run the setup code chunk below in your Quarto notebook to create the data frame we will use in this lesson.

```{r}
#| label: setup
#| code-fold: true

library(tidyverse)
library(vdemdata)

vdem2022 <- vdem |>
  filter(year == 2022) |>
  select(
    country = country_name,
    polyarchy = v2x_polyarchy,
    women_empowerment = v2x_gender,
    polarization = v2cacamps,
    regime = v2x_regime,
    region = e_regionpol_6C
  ) |>
  mutate(
    region = case_match(region,
      1 ~ "Eastern Europe",
      2 ~ "Latin America",
      3 ~ "Middle East",
      4 ~ "Africa",
      5 ~ "The West",
      6 ~ "Asia"),
    regime = case_match(regime,
      0 ~ "Closed Autocracy",
      1 ~ "Electoral Autocracy",
      2 ~ "Electoral Democracy",
      3 ~ "Liberal Democracy")
  )
```
:::

## Overview

In this module we explore how to work with **categorical data** — variables that place observations into groups rather than measuring them on a continuous scale. We begin by discussing the different types of data you will encounter in political and social science research, and why those distinctions matter. Then, using democracy indicators from V-Dem, we visualize the distribution of categorical variables and examine how they vary across groups.

{{< video https://www.youtube.com/watch?v=7NhNeADL8fA title='Exploring Categorical Data' >}}

## What Kind of Data Do We Have?

Before summarizing or modeling a dataset, it helps to step back and think about what kind of data we are actually dealing with. This shapes the types of questions we can ask, the summaries we can produce, and the visualizations we can make.

One of the most fundamental distinctions is between **categorical** and **numerical** variables. **Categorical variables** place observations into groups — like types of political regimes or regions of the world. **Numerical variables** are measured on a scale — like GDP per capita or a democracy score.

Among numerical variables, we distinguish between **discrete** variables (whole-number counts, like the number of protest events) and **continuous** variables (any value in a range, like income or temperature). We will focus on continuous variables in the next module.

It also matters how the data were collected. Are we working with a **cross-sectional** snapshot of one point in time? Or a **time series** tracking change over time? Most of the datasets we use in this course are cross-sectional, but V-Dem is a panel dataset — it has both a country and a year dimension.

Finally, not all categories are created equal. **Nominal** categorical variables have no inherent order — like world region or country name. **Ordinal** categorical variables have a meaningful order — like V-Dem's regime classification, which runs from closed autocracy to liberal democracy. Understanding this distinction matters when we choose how to visualize and interpret our data.

::: {.callout-warning icon=false}
## Your Turn

Classify the following variables:

- Is a country a democracy? (yes/no)
- Polity score (ranges from −10 to 10)
- V-Dem Polyarchy index (0 to 1)
- V-Dem Regimes of the World (closed autocracy, electoral autocracy, etc.)
- Number of protest events in a year
- Protest type (sit-in, march, strike, etc.)

For each: Is it categorical or numerical? If categorical, is it nominal or ordinal? If numerical, is it discrete or continuous?
:::

## Exploring Categorical Data

Let's look at a real categorical variable: V-Dem's **Regimes of the World** classification, which sorts countries into four types:

- Closed Autocracy
- Electoral Autocracy
- Electoral Democracy
- Liberal Democracy

These categories follow a meaningful order from least to most democratic, making this an **ordinal categorical** variable. We start by glimpsing the data:

```{r}
glimpse(vdem2022)
```

A quick frequency table with `count()` shows how many countries fall into each category:

```{r}
vdem2022 |>
  count(regime)
```

A bar chart gives us the same information visually. We use `geom_bar()`, which automatically counts observations for each category — you only need to specify the x aesthetic:

```{r}
vdem2022 |>
  ggplot(aes(x = regime)) +
  geom_bar(fill = "steelblue") +
  labs(
    x = "Regime Type",
    y = "Frequency",
    title = "Regimes of the World in 2022",
    caption = "Source: V-Dem Institute"
  ) +
  theme_minimal()
```

Electoral democracies are the most common regime type in 2022, followed closely by electoral autocracies.

::: {.callout-note}
`geom_bar()` and `geom_col()` both make bar charts, but they work differently. `geom_bar()` counts observations automatically — you only need to specify `x`. `geom_col()` requires you to supply both `x` and `y`, where `y` is a column of pre-computed values. Use `geom_bar()` when working directly with raw data; use `geom_col()` when you have already summarized the data yourself.
:::

::: {.callout-warning icon=false}
## Your Turn

- Filter the data to a different year and visualize the distribution of regime types.
- Update the plot title to reflect the year you chose.
- What differences do you notice compared to 2022?
:::

## Regimes by Region

Now let's break the distribution down by world region. We map `region` to the x-axis and use the `fill` aesthetic to color bars by regime type:

```{r}
vdem2022 |>
  ggplot(aes(x = region, fill = regime)) +
  geom_bar() +
  theme_minimal() +
  labs(
    x = "Region",
    y = "Frequency",
    title = "Regimes of the World by Region in 2022",
    caption = "Source: V-Dem Institute",
    fill = "Regime"
  ) +
  scale_fill_viridis_d(option = "inferno", end = .8)
```

One challenge immediately appears: Africa has far more countries than Eastern Europe or the Middle East. Even if the proportions were identical, Africa's bars would simply be taller. To make a fair comparison across regions, we can switch from counts to proportions using `position = "fill"`:

```{r}
vdem2022 |>
  ggplot(aes(x = region, fill = regime)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(
    x = "Region",
    y = "Proportion",
    title = "Regimes of the World by Region in 2022",
    caption = "Source: V-Dem Institute",
    fill = "Regime"
  ) +
  scale_fill_viridis_d(option = "inferno", end = .8)
```

Now we can compare across regions on equal footing. The West stands out for the prevalence of liberal democracies; the Middle East and Africa have the highest concentration of closed and electoral autocracies.

Note what kinds of variables we are combining here. **Region** is a **nominal** categorical variable — there is no meaningful order to Asia, Africa, or Latin America. **Regime type** is an **ordinal** categorical variable — its categories run from least to most democratic. Recognizing this distinction helps us think about how to display and interpret the data.

::: {.callout-warning icon=false}
## Your Turn

- Switch the x-axis from `region` to `regime` and the `fill` to `region`. What does this version of the plot emphasize?
- Try `position = "dodge"` instead of `position = "fill"`. What are the trade-offs between the stacked proportional chart and the side-by-side chart?
- Use `filter()` to restrict the plot to a single region. What do you find?
:::
