---
title: Logistic Regression
subtitle: Classification
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---


```{r}
#| label: packages
#| include: false
library(tidyverse)
library(peacesciencer)
library(broom)
```

# Binary Outcomes

## Binary Outcomes

<br>

- So far we have looked at continuous or numerical outcomes (response variables)
- We are often also interested in outcome variables that are binary (Yes/No, or 1/0)
    - Did violence happen, or not?
    - Classification: is this email spam?

## Example: Conflict Onset

<br>

::: {.incremental}
- Did a civil war begin in a given country in a given year? (yes/no)
- Predictors: wealth, democracy, terrain, ethnic diversity, etc. 
- Seminal work by Fearon and Laitin (2003) 
- We can use logistic regression to model this binary outcome
:::

## Modeling

<br>

::: {.incremental}
- We can treat each outcome (conflict onset) as successes and failures arising from separate Bernoulli trials
- **Bernoulli trial**: a random experiment with exactly two possible outcomes, "success" and "failure", in which the probability of success is the same every time the experiment is conducted
- Success is usually coded as 1, failure as 0
- So ironically, conflict onset is a "success" in this context
:::

## Modeling

<br>

Each Bernoulli trial can have a separate probability of success

<br>

$$ y_i ∼ Bern(p_i) $$

<br>

. . .

We can then use the predictor variables to model that probability of success, $p_i$

## Why Not Linear Regression?

```{r}
#| label: lpm-problem

set.seed(42)
n <- 200
x <- rnorm(n, mean = 0, sd = 2)
p_true <- plogis(0.8 * x)
y <- rbinom(n, 1, p_true)
df_lpm <- tibble(x = x, y = y)

lpm_fit <- lm(y ~ x, data = df_lpm)
df_lpm <- df_lpm |> mutate(y_hat = predict(lpm_fit))

ggplot(df_lpm, aes(x = x, y = y)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = y_hat), color = "firebrick", linewidth = 1) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "gray50") +
  labs(title = "Linear Probability Model",
       subtitle = "Predicted values can fall outside [0, 1]",
       x = "Predictor", y = "Outcome (0/1)") +
  theme_minimal()
```

## Why Not Linear Regression?

::: {.incremental}
- With a binary outcome (0/1), OLS can predict values **outside [0, 1]**
- Predicted probabilities below 0 or above 1 are not meaningful
- OLS assumes a linear relationship between predictors and the expected value of (Y)
- For binary outcomes, the relationship between predictors and probabilities is usually not linear over the full range
- We need a model that lets the mean of (Y) depend on predictors in a way appropriate to the outcome
:::

## Modeling

<br>

- We can't really use a linear model for $p_i$ (since $p_i$ must be between 0 and 1 and the relationship is not linear) but we can transform the linear model to have the appropriate range
- So instead of OLS, we use a family of models called **generalized linear models (GLMs)**
- Logistic regression is a very common example

## GLMs

<br>

All GLMs have the following three characteristics:

::: {.incremental}
- An outcome distribution describing a generative model for the outcome variable (e.g. Bernoulli, Poisson, normal)
- A linear predictor:
$$\eta = \beta_0 + \beta_1 X_1 + \cdots + \beta_k X_k$$
- A link function that connects the expected value of the outcome to the linear predictor: $$g(\mu_i) = \eta_i$$
:::

## The Sigmoid Function

<br>

- The **logistic (sigmoid) function** maps any real number to (0, 1):

$$p = \frac{\exp(\eta)}{1 + \exp(\eta)}$$

- As $\eta \to -\infty$, $p \to 0$; as $\eta \to +\infty$, $p \to 1$
- This S-shaped curve is the inverse of the logit function

## The Sigmoid Function

```{r}
#| label: sigmoid

tibble(eta = seq(-6, 6, length.out = 300)) |>
  mutate(p = exp(eta) / (1 + exp(eta))) |>
  ggplot(aes(x = eta, y = p)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "gray60") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "gray40") +
  labs(title = "The Sigmoid (Logistic) Function",
       x = expression(eta ~ "= linear predictor"),
       y = expression(p ~ "= probability")) +
  theme_minimal()
```

## Logistic Regression

<br>

- Logistic regression is a GLM used to model a binary categorical outcome (0 or 1)
- In logistic regression, the link function that connects $\eta_i$ to $p_i$ is the **logit function**
- **Logit function:** For $0\le p \le 1$

$$logit(p) = \log\left(\frac{p}{1-p}\right)$$

## Logit Function


```{r}
#| label: logit_function

d <- tibble(p = seq(0.001, 0.999, length.out = 1000)) |>
  mutate(logit_p = log(p/(1-p)))

ggplot(d, aes(x = p, y = logit_p)) + 
  geom_line() + 
  xlim(0,1) + 
  ylab("logit(p)") +
  labs(title = "logit(p) vs. p")
```


## Logistic Regression Model

<br>

::: {.incremental}
- $y_i \sim \text{Bern}(p_i)$
- $\eta_i = \beta_0+ \beta_1 x_{1,i} + \cdots + \beta_n x_{n,i}$
- $\text{logit}(p_i) = \eta_i$
:::

## Logistic Regression Model

<br>

- $\text{logit}(p_i) = \eta_i = \beta_0+ \beta_1 x_{1,i} + \cdots + \beta_n x_{n,i}$
- Now take inverse logit to get $p$

$$p_i = \frac{\exp(\beta_0+\beta_1 x_{1,i} + \cdots + \beta_k x_{k,i})}{1+\exp(\beta_0+\beta_1 x_{1,i} + \cdots + \beta_k x_{k,i})}$$

# Analyzing Conflict Onset

## The `peacesciencer` Package

- The `peacesciencer` package provides a number of datasets and functions for analyzing conflict and peace
- Provides data from a number of important datasets in the field of conflict studies, e.g.
  - Correlates of War (CoW) project
  - Uppsala Conflict Data Program (UCDP)
  - Militarized Interstate Dispute (MID) dataset
- Provides functions for analyzing conflict and adding control variables to the dataset

## Using the `peacesciencer` Package

<br>

```{r}
#| label: use_peacesciencer
#| echo: true

conflict_df <- create_stateyears(system = 'gw') |>
  filter(year %in% c(1946:1999)) |>
  add_ucdp_acd(type=c("intrastate"), only_wars = FALSE) |>
  add_democracy() |>
  add_creg_fractionalization() |>
  add_sdp_gdp() |>
  add_rugged_terrain()

glimpse(conflict_df)
```

## Running a Logistic Regression

<br>

- Implementation is similar to `lm()` for linear regression
- We use `glm()` with `family = "binomial"` to fit a logistic model
- The `~` formula syntax is the same as in `lm()`

## Bivariate Logistic Regression

<br>

```{r}
#| label: bivariate_model
#| echo: true

conflict_model <- glm(ucdponset ~ wbgdppc2011est,
                  data= conflict_df,
                  family = "binomial")

summary(conflict_model)
```

```{r}
#| label: helpers
#| include: false

b0 <- round(coef(conflict_model)[["(Intercept)"]], 2)
b1 <- round(coef(conflict_model)[["wbgdppc2011est"]], 2)
b1_fmt <- ifelse(b1 >= 0, paste0("+ ", b1), paste0("- ", abs(b1)))
or_b1 <- round(exp(b1), 2)
pct_change_b1 <- round((1 - exp(b1)) * 100, 1)
logodds_9 <- round(b0 + b1 * 9, 2)
odds_9 <- round(exp(logodds_9), 3)
odds_9_plus_1 <- round(1 + odds_9, 3)
prob_9 <- round(odds_9 / (1 + odds_9), 4)
```

## Interpreting the Results

<br>


$$\log\left(\frac{p}{1-p}\right) = `r b0` `r b1_fmt` \times \text{logGDPpc}$$

## Interpreting the Results

<br> 

- For a quick interpretation of the coefficients, we can exponentiate them
- The exponentiated coefficient is the odds ratio
- For each one-unit increase in the independent variable, the odds of the outcome occurring increase (or decrease) by a factor of the exponentiated coefficient

## Interpreting the Results {.smaller}

<br>

$$\log\left(\frac{p}{1-p}\right) = `r b0` `r b1_fmt` \times \text{logGDPpc}$$

<br>

For each one unit increase in log GDP per capita, the odds of the outcome occurring are multiplied by approximately `r or_b1`, assuming other variables in the model are held constant.

<br>

This means that an increase in GDP per capita is associated with a **decrease** in the odds of the outcome occurring. The odds of the outcome decrease by about `r pct_change_b1`% for each unit increase in GDP per capita (on average).

## Your Turn! {.smaller}

<br>

- Run a bivariate logistic regression using ucdp onset as the outcome variable
- First replicate the results using GDP per capita as the predictor
- Now try a different predictor
- Interpret the results
  - What is the average effect of the predictor on conflict onset? 
  - How do you interpret that effect in terms of the odds of conflict onset?
  
```{r}
#| label: time1

library(countdown)

countdown(minutes = 10, 
          id = "timer1", 
          bottom = "5%", 
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

## Calculating Predicted Probabilities {.scrollable}

<br>

Probability of conflict onset for a country with a log per capita GDP of 9 (about $8,000):

$$\log\left(\frac{p}{1-p}\right) = `r b0` `r b1_fmt` \times 9$$
$$\log\left(\frac{p}{1-p}\right) = `r logodds_9`$$

$$\frac{p}{1-p} = \exp(`r logodds_9`)$$

$$\frac{p}{1-p} = `r odds_9`$$

$$p = `r odds_9` \times (1 - p)$$
$$p = `r odds_9` - `r odds_9`p$$

$$`r odds_9_plus_1`p = `r odds_9`$$
$$p = `r odds_9` / `r odds_9_plus_1`$$
$$p = `r prob_9`$$

## Using `marginaleffects` 

<br>

```{r}
#| label: marginal_effects
#| echo: true
#| eval: false

library(marginaleffects)

# select some countries for a given year
selected_countries <- conflict_df |>
  filter(
    gw_name %in% c("United States of America", "Venezuela", "Rwanda"),
    year == 1999)

# calculate margins for the subset
marg_effects <- predictions(conflict_model, newdata = selected_countries)

tidy(marg_effects) |>
  select(gw_name, estimate, p.value, conf.low, conf.high)
```

## Using `marginaleffects`

<br>

```{r}
#| label: marginal_effects2

library(marginaleffects)

selected_countries <- conflict_df |>
  filter(
    gw_name %in% c("United States of America", "Venezuela", "Rwanda"),
    year == 1999)

marg_effects <- predictions(conflict_model, newdata = selected_countries)

tidy(marg_effects) |>
  select(gw_name, estimate, p.value, conf.low, conf.high)
```


## Your Turn!

<br>

- Select your favorite three countries and a recent year
- Calculate the predicted probability of conflict onset for that year using the marginal effects package
- If you have time, try to do the calculation by hand as well

```{r}
#| label: time2

countdown(minutes = 10, 
          id = "timer2", 
          bottom = "10%", 
          right = "10%")
```

