---
title: Sampling and Uncertainty
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
filters:
  - shinylive
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## Sampling

<br>

::: incremental
- **Sampling** the act of selecting a subset of individuals, items, or data points from a larger population to estimate characteristics or metrics of the entire population
- Versus a **census**, which involves gathering information on every individual in the population
- Why would you want to use a sample?
:::

# Sampling Activity

## What proportion of all milk chocolate M&Ms are blue?

<br>

-   M&Ms has a precise distribution of colors that it produces in its factories
-   M&Ms are sorted into bags in factories in a fairly random process
-   Each bag represents a **sample** from the full **population** of M&Ms

## Activity

<br>

- Use the virtual bag sampler on the next slide
- Click **Open a New Bag** three or four times — each click is one sample from the M&M population
- Record your proportion estimates (the app tracks them for you)
- Add your estimates to the [class Google Sheet](https://docs.google.com/spreadsheets/d/136wGKZOnwOdo3O-4bUfF_TWKBCAWQUCyiVKSY4HyYAw/edit?usp=sharing) — we will analyze the class data together

## Virtual M&M Bag Sampler

```{shinylive-r}
#| standalone: true
#| viewerHeight: 540

library(shiny)

true_prop_blue <- 0.24
mm_probs <- c(blue = 0.24, orange = 0.20, green = 0.16,
              yellow = 0.14, red = 0.13, brown = 0.13)
bag_size <- 56

ui <- fluidPage(
  fluidRow(
    column(4,
      wellPanel(
        actionButton("open_bag", "Open a New Bag",
                     class = "btn-primary btn-lg",
                     style = "width:100%; margin-bottom:12px;"),
        conditionalPanel("input.open_bag > 0",
          h4("Your Estimates"),
          tableOutput("estimates_table"),
          hr(),
          strong(textOutput("running_mean"))
        )
      )
    ),
    column(8,
      plotOutput("bag_plot", height = "270px"),
      conditionalPanel("input.open_bag > 1",
        plotOutput("hist_plot", height = "190px")
      )
    )
  )
)

server <- function(input, output, session) {

  estimates   <- reactiveVal(numeric(0))
  current_bag <- reactiveVal(NULL)

  observeEvent(input$open_bag, {
    bag <- sample(names(mm_probs), size = bag_size, replace = TRUE,
                  prob = mm_probs)
    current_bag(bag)
    estimates(c(estimates(), mean(bag == "blue")))
  })

  output$bag_plot <- renderPlot({
    req(current_bag())
    colors  <- current_bag()
    n       <- length(colors)
    ncols   <- 8
    nrows   <- ceiling(n / ncols)
    x <- rep(1:ncols, length.out = n)
    y <- rep(nrows:1, each = ncols)[1:n]
    prop_blue <- mean(colors == "blue")

    par(mar = c(1, 1, 3, 1))
    plot(x, y,
         xlim = c(0.5, ncols + 0.5), ylim = c(0.5, nrows + 0.5),
         type = "n", axes = FALSE, xlab = "", ylab = "",
         main = sprintf("Bag #%d  —  Proportion Blue: %.3f  (%d / %d)",
                        input$open_bag, prop_blue,
                        sum(colors == "blue"), n))
    for (i in seq_along(colors)) {
      symbols(x[i], y[i], circles = 0.35,
              fg = if (colors[i] == "blue") "navy" else "gray60",
              bg = colors[i],
              add = TRUE, inches = FALSE,
              lwd = if (colors[i] == "blue") 2.5 else 1)
    }
  })

  output$estimates_table <- renderTable({
    req(length(estimates()) > 0)
    e <- estimates()
    data.frame(Bag = paste0("#", seq_along(e)),
               `Prop. Blue` = round(e, 3), check.names = FALSE)
  }, striped = TRUE, bordered = TRUE, spacing = "s")

  output$running_mean <- renderText({
    e <- estimates()
    req(length(e) > 0)
    sprintf("Mean of %d bags: %.3f", length(e), mean(e))
  })

  output$hist_plot <- renderPlot({
    req(length(estimates()) > 1)
    e <- estimates()
    par(mar = c(4, 4, 3, 1))
    hist(e,
         breaks = seq(0, 0.65, by = 0.05),
         col = "steelblue3", border = "white",
         xlim = c(0, 0.6),
         xlab = "Proportion Blue", ylab = "Count",
         main = "Your sampling distribution so far",
         xaxt = "n")
    axis(1, at = seq(0, 0.6, by = 0.1),
         labels = paste0(seq(0, 60, by = 10), "%"))
    abline(v = true_prop_blue, col = "red", lty = 2, lwd = 2)
    text(true_prop_blue + 0.02, par("usr")[4] * 0.85,
         "True: 24%", col = "red", adj = 0, cex = 0.9)
  })
}

shinyApp(ui, server)
```

## Let's Analyze the Data

<br>

We will use the `googlesheets4` package to pull the data into R so be sure to install it.

```{r}
#| label: load_mm_data
#| echo: true
#| eval: false

#install.packages("googlesheets4")

library(tidyverse)
library(googlesheets4)

gs4_deauth() # to signify no authorization required

mnm_data <- read_sheet("https://docs.google.com/spreadsheets/d/136wGKZOnwOdo3O-4bUfF_TWKBCAWQUCyiVKSY4HyYAw/edit#gid=0")

glimpse(mnm_data)
```


## Calculate Some Summary Stats

<br>

```{r}
#| label: summary_stats
#| echo: true
#| eval: false

mnm_data |>
  summarize(
    mean_blue = mean(proportion_blue),
    median_blue = median(proportion_blue),
    sd_blue = sd(proportion_blue)
  )
```

## Now Let's Make a Histogram 

<br>

```{r}
#| label: histogram
#| echo: true
#| eval: false

ggplot(mnm_data, aes(x = proportion_blue)) +
  geom_histogram(fill = "steelblue") +
  labs(
    title = "Percentage of Blue M&Ms",
    x = "Proportion Blue",
    y = "Count"
  )
```

## Discussion

<br>

-   What is the histogram/distribution showing?
-   Why do some bag of M&Ms have proportions of blues that are higher and lower than the number you gave above?
-   How do our estimates relate to the actual percentage of blue M&Ms manufactured?
-   What do you think would happen if we kept adding more samples to the distribution?

## What did we just do?

<br>

::: incremental
-   We wanted to say something about the **population** of M&Ms
-   The **parameter** we care about is the proportion of M&Ms that are blue
-   It would be impossible to conduct a **census** and to calculate the parameter
-   We took a **sample** from the population and calculated a **sample statistic**
-   **statistical inference:** act of making a guess about a **population** using information from a **sample**
:::

## What did we just do?

<br>

::: incremental
-   We completed this task many times
-   This produced a **sampling distribution** of our **estimates**
-   There is a distribution of estimates because of **sampling variability** 
-   Due to random chance, one estimate from one sample can differ from another
-   These are foundational ideas for statistical inference that we are going to keep building on
:::

# Zooming Out

## Target Population

<br>

In data analysis, we are usually interested in saying something about a **Target Population.**

::: incremental
-   What proportion of adult Russians support the war in Ukraine?
    -   Target population: adult Russians (age 18+)
-   How many US college students check social media during their classes?
    -   Target population: US college students
-   What percentage of M&Ms are blue?
    -   Target population: all of the M&Ms
:::

## Sample

<br>

In many instances, we have a **Sample**

-   We cannot talk to every Russian
-   We cannot talk to all college students
-   We cannot count all of the M&Ms

## Parameters vs Statistics

<br>

::: incremental
-   The **parameter** is the value of a calculation for the entire target population
-   The **statistic** is what we calculate on our sample
    -   We calculate a statistic in order to say something about the parameter
:::

## Inference

<br>

::: incremental
-   **Inference**--The act of "making a guess" about some unknown
-   **Statistical inference**--Making a good guess about a population from a sample
-   **Causal inference**--Did X cause Y? [topic for later classes]
:::

# Uncertainty

## {.smaller}

On December 19, 2014, the front page of Spanish national newspaper El
País read *"Catalan public opinion swings toward 'no' for independence, says survey"*.^[Alberto Cairo. [The truthful art: Data, charts, and maps for communication](http://www.thefunctionalart.com/p/the-truthful-art-book.html). New Riders, 2016.]

```{r}
library(tidyverse)
library(lubridate)
library(scales)
catalan <- tibble(
  response = c("No", "Yes", "No answer"),
  rate     = c(45.3, 44.5, 10.2)
) |>
  mutate(response = fct_relevel(response, "No", "Yes", "No answer"))
```

```{r}
#| label: catalan-misleading
ggplot(catalan, aes(y = fct_rev(response), x = rate, fill = response)) +
  geom_col(width = .75) +
  scale_fill_manual(values = c("#5C8AA9", "#9D303A", "gray")) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  theme_minimal(base_size = 16) +
  labs(
    title = "Do you want Catalonia\nto become an independent state?",
    caption = "Margin of error: +/-2.95% at 95% confidence level",
    x = NULL, y = NULL) +
  theme(legend.position = "none")
```

## {.smaller}

The probability of the tiny difference between the 'No' and 'Yes' being just due to random chance is very high.^[Alberto Cairo. ["Uncertainty and Graphicacy"](https://ec.europa.eu/eurostat/cros/powerfromstatistics/OR/PfS-OutlookReport-Cairo.pdf), 2017.]

```{r}
catalan <- catalan |>
  mutate(
    low = rate - 2.95,
    high = rate + 2.95
  )
ggplot(catalan, aes(y = fct_rev(response), x = rate, color = response, group = response)) +
  geom_segment(aes(x = low, xend = high, 
                   y = fct_rev(response), yend = fct_rev(response)),
               linewidth = 0.8, color = "black") +
  geom_point(size = 3) +
  scale_color_manual(values = c("#5C8AA9", "#9D303A", "gray")) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  guides(color = "none") +
  theme_minimal(base_size = 16) +
  labs(
    title = "Do you want Catalonia to become an independent state?",
    x = NULL, y = NULL
  )
```

## Characterizing Uncertainty

<br>

- We know from previous section that even unbiased procedures do not get the "right" answer every time
- We also know that our estimates might vary from sample to sample due to random chance
- Therefore we want to report on our estimate and our level of uncertainty


## Characterizing Uncertainty

<br>

- With M&Ms, we knew the population parameter
- In real life, we do not!
- We want to generate an estimate *and* characterize our uncertainty with a *range* of possible estimates
    
## Solution: Create a Confidence Interval

<br>

- A plausible range of values for the population parameter is a confidence interval.

. . .

- 95 percent confidence interval is standard
    - We are 95% confident that the parameter value falls within the range given by the confidence interval

## Ways to Estimate

<br>

- Take advantage of Central Limit Theorem to estimate using math
- Use simulation, bootstrapping 

## With Math...

$$CI = \bar{x} \pm Z \left( \frac{\sigma}{\sqrt{n}} \right)$$

- $\bar{x}$ is the sample mean,
- $Z$ is the Z-score corresponding to the desired level of confidence
- $\sigma$ is the population standard deviation, and 
- $n$ is the sample size

## 

<br>

This part here represents the standard error: 

$$\left( \frac{\sigma}{\sqrt{n}} \right)$$

- Standard deviation of the sampling distribution
- Characterizes the spread of the sampling distribution
- The bigger this is the bigger the CIs are going to be

## Central Limit Theorem

$$CI = \bar{x} \pm Z \left( \frac{\sigma}{\sqrt{n}} \right)$$

- This way of doing things depends on the Central Limit Theorem
- As sample size gets bigger, the spread of the sampling distribution gets narrower
- The shape of the sampling distributions becomes more normally distributed

## 

<br>

$$CI = \bar{x} \pm Z \left( \frac{\sigma}{\sqrt{n}} \right)$$

This is therefore a **parametric** method of calculating the CI. It depends on assumptions about the normality of the distribution.  

## Bootstrapping

<br>

::: incremental
- Pulling oneself up from their bootstraps ... 
- Use the data we have to estimate the sampling distribution
- We call this the *bootstrap* distribution
- This is a **nonparametric** method
- It does not depend on assumptions about normality
:::

## Bootstrap Process {.smaller}

<br>

1. Take a bootstrap sample - a random sample taken **with replacement** from the original sample, of the **same size** as the original sample

. . .

2. Calculate the bootstrap statistic - a statistic such as mean, median, proportion, slope, etc. computed on the bootstrap samples

. . .

3. Repeat steps (1) and (2) many times to create a bootstrap distribution - a distribution of bootstrap statistics

. . .

4. Calculate the bounds of the XX% confidence interval as the middle XX% 
of the bootstrap distribution (usually 95 percent confidence interval)


## Russia

<br>

What Proportion of Russians believe their country interfered in the 2016 presidential elections in the US?

- Pew Research survey
- 506 subjects
- Data available in the `openintro` package

## 

<br>

For this example, we will use data from the Open Intro package. Install that package before running this code chunk.

<br>

```{r}
#| echo: true

#install.packages("openintro")
library(openintro)

glimpse(russian_influence_on_us_election_2016)
```


## 

<br>

Let's use `mutate()` to recode the qualitative variable as a numeric one...

<br>

```{r}
#| echo: true

russiaData <- russian_influence_on_us_election_2016 |> 
  mutate(try_influence = ifelse(influence_2016 == "Did try", 1, 0))
```

##

<br>

Now let's calculate the mean and standard deviation of the `try_influence` variable... 

<br>

```{r}
#| echo: true

russiaData |>
  summarize( 
          mean = mean(try_influence),
          sd = sd(try_influence)
  )
```

## 

<br>

And finally let's draw a bar plot...

<br>

```{r}
#| echo: true
#| eval: false

ggplot(russiaData, aes(x = try_influence)) +
  geom_bar(fill = "steelblue", width = .75) +
  labs(
    title = "Did Russia try to influence the U.S. election?",
    x = "0 = 'No', 1 = 'Yes'",
    y = "Frequncy"
  ) +
  theme_minimal()
```

## 

<br>

```{r}
ggplot(russiaData, aes(x = try_influence)) +
  geom_bar(fill = "steelblue", width = .75) +
  labs(
    title = "Did Russia try to influence the U.S. election?",
    x = "0 = 'No', 1 = 'Yes'",
    y = "Frequncy"
  ) +
  theme_minimal()
```

## Bootstrap with `tidymodels`

<br>

Install `tidymodels` before running this code chunk... 

```{r}
#| echo: true
#| code-line-numbers: "|1-2|4|5|6-7|8-9|10-11|13"

#install.packages("tidymodels")
library(tidymodels)

set.seed(66)
boot_dist <- russiaData |>
  # specify the variable of interest
  specify(response = try_influence) |>
  # generate 10000 bootstrap samples
  generate(reps = 10000, type = "bootstrap") |>
  # calculate the mean of each bootstrap sample
  calculate(stat = "mean")

glimpse(boot_dist)
```

## 

<br>

Calculate the mean of the bootstrap distribution (of the means of the individual draws)...

<br>

```{r}
#| echo: true

boot_dist |> summarize(mean = mean(stat))
```

## 

<br>

Calculate the confidence interval. A 95% confidence interval is bounded by the middle 95% of the bootstrap distribution.

<br>

```{r}
#| echo: true 

boot_dist |>
  summarize(lower = quantile(stat, 0.025),
            upper = quantile(stat, 0.975))
```

## Confidence Interval with `infer`

<br>

```{r}
#| label: infer-get-ci
#| echo: true

ci <- boot_dist |> get_ci(level = 0.95)

ci
```

## Visualize with `infer`

<br>

```{r}
#| label: infer-visualize-code
#| echo: true
#| eval: false
#| code-line-numbers: "2-3"

boot_dist |>
  visualize() +
  shade_ci(ci, color = "red", fill = NULL) +
  labs(
    title = "Bootstrap Distribution with 95% CI",
    x = "Mean",
    y = "Count"
  ) +
  theme_minimal()
```

## Visualize with `infer`

<br>

```{r}
#| label: infer-visualize
#| code-fold: true

boot_dist |>
  visualize() +
  shade_ci(ci, color = "red", fill = NULL) +
  labs(
    title = "Bootstrap Distribution with 95% CI",
    x = "Mean",
    y = "Count"
  ) +
  theme_minimal()
```

## Interpret the confidence interval {.smaller}

<br>

The 95% confidence interval was calculated as (`r round(ci$lower_ci, 3)`, `r round(ci$upper_ci, 3)`). Which of the following is the correct interpretation of this interval?

<br>

**(a)** 95% of the time the percentage of Russians who believe that Russia interfered in the 2016 US elections is between `r round(ci$lower_ci, 3)` and `r round(ci$upper_ci, 3)`.

**(b)** 95% of all Russians believe that the chance Russia interfered in the 2016 US elections is between `r round(ci$lower_ci, 3)` and `r round(ci$upper_ci, 3)`.

**(c)** We are 95% confident that the proportion of Russians who believe that Russia interfered in the 2016 US election is between `r round(ci$lower_ci, 3)` and `r round(ci$upper_ci, 3)`.

**(d)** We are 95% confident that the proportion of Russians who supported interfering in the 2016 US elections is between `r round(ci$lower_ci, 3)` and `r round(ci$upper_ci, 3)`.

## Your Turn! {.smaller}

<br>

- Change the `reps` argument in the `generate()` function to something way smaller, like 100. What happens to your estimates?
- Try progressively higher numbers. What happens to your estimates as the number of reps increases?

## Bias vs Precision

- A procedure is *unbiased* if it generates the "right" answer, on average
- Precision refers to variability: procedures with less sampling variability will be more precise
    - all else equal, a greater sample size will increase precision
- When we increase the sample size (number of reps), we increase precision
- As a result our confidence interval will be narrower

## Why did we do these simulations?

<br>

- They provide a foundation for statistical inference and for characterizing *uncertainty* in our estimates
- The best **research designs** will try to maximize or achieve good balance on bias vs precision