---
title: Multiple Regression
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## Load packages 

<br>

```{r}
#| label: packages
#| echo: true

library(tidyverse)
library(vdemdata)
library(broom)
```


## Load VDEM Data {.smaller}

<br>

```{r}
#| label: get_data
#| echo: true

modelData <- vdem |>
  filter(year == 2006) |> 
  select(country_name, 
         libdem = v2x_libdem, 
         wealth = e_gdppc, 
         oil_rents = e_total_oil_income_pc,
         polarization = v2cacamps, 
         corruption = v2x_corr, 
         judicial_review = v2jureview_ord, 
         region = e_regionpol_6C, 
         regime = v2x_regime) |> 
  mutate(log_wealth = log(wealth), 
         region = factor(
           region,
           labels=c("Eastern Europe", 
             "Latin America", 
             "MENA", 
             "SSAfrica", 
             "Western Europe and North America", 
             "Asia and Pacific"))
          )

glimpse(modelData)
```

## Linear Model with Multiple Predictors

<br>

- Previously, we were interested in GDP per capita as a predictor of democracy
- Now, let's consider another predictor: polarization (also measured by V-Dem)

## Polarization Measure in the USA

```{r}
vdem |> 
  filter(country_name == "United States of America") |>
  filter(year > 1900) |> 
  ggplot(aes(y= v2cacamps, x = year)) +
  geom_line(color = "dodgerblue", linewidth = 1) +
  labs(title = "Polarization in the USA", 
       y = "Polarization", x = "Year") +
  theme_minimal()
```

## Polarization and Democracy in 2006

```{r}
#| echo: false
ggplot(modelData, aes(x = polarization, y = libdem)) +
  geom_point() +
  geom_smooth(method = "lm", color = "#E48957", se = FALSE) +
  labs(x = "Polarization", y = "Liberal Democracy Index") +
  theme_minimal()
```

## Model with One Predictor

<br>

```{r}
#| label: model1
#| echo: true
  
model1 <- lm(libdem ~ polarization, data = modelData) 

summary(model1)
```
<br>

| How do we interpret the intercept and the slope estimate?
| Significance? 

## Model with Two Predictors

<br>

```{r}
#| label: model2
#| echo: true

model2 <- lm(libdem ~ polarization + log_wealth, data = modelData) 

summary(model2)
```

## Model with Two Predictors

<br>

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 


## Model with Two Predictors 

<br>

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 

$a$ is the predicted level of Y when BOTH GDP per capita and polarization are equal to 0

## Model with Two Predictors 

<br>

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 

- $b_1$ is the impact of a 1-unit change in polarization on the predicted level of Y, **holding GDP per capita fixed** (all else equal)
- The relationship between polarization and democracy, controlling for wealth

## Model with Two Predictors 

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 

- $b_2$ is the impact of a 1-unit change in GDP per capita on the predicted level of Y, **holding polarization fixed** (all else equal)
- The relationship between wealth and democracy, controlling for polarization

## Model with Two Predictors

<br>

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 


- OLS is searching for combination of $a$, $b_1$, and $b_2$ that minimize sum of squared residuals
- Same logic as model with one predictor, just more complicated

## Model with Three Predictors 

<br>

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc + b_3*OilRents$$ 

## Model with Three Predictors 

<br>

```{r}
#| label: model3
#| echo: true


model3 <- lm(libdem ~ polarization + log_wealth + oil_rents, data = modelData) 

summary(model3)
```

## Model with Three Predictors {.smaller}

<br>

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc + b_3*OilRents$$ 

$$\hat{Y_i} = a + -.05*Polarization + .13*GDPpc + .00004*OilRents$$ 
<br>

| $b_3$ is the impact of a 1-unit change in oil revenues per capita on the predicted level of Y, **holding polarization and GDP fixed** (all else equal)

## Making Predictions

<br>

Use `predict()` to get predicted values for specific cases:

```{r}
#| label: predict
#| echo: true

# predict democracy for three hypothetical countries
new_data <- tibble(
  polarization = c(-1, 0, 1),
  log_wealth   = c(7, 9, 11)
)

predict(model2, newdata = new_data)
```

<br>

Each value is the predicted liberal democracy score for a country with those characteristics, holding other factors at the specified values.

## Your Turn! {.smaller}

<br>

- In the last session, you examined levels of democracy and corruption
- Now, let's fit a multiple regression model predicting corruption with two predictors: democracy (libdem) and polarization (polarization)
- Then, interpret the coefficients
- Estimate a second multiple regression model that adds in GDP per capita (log_wealth)
- Interpret the coefficients
- What happens to the coefficients of the other variables when we add GDP per capita to the model?
- Try additional predictors if there is time

```{r}
#| label: time1

library(countdown)

countdown(minutes = 10, 
          id = "timer1", 
          bottom = "10%", 
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

## Two Predictors

<br>

```{r}
#| echo: true

corruption1 <- lm(corruption ~ libdem + polarization , data = modelData) 

summary(corruption1)
```

## Adding GDP per capita

<br>

```{r}
#| echo: true

corruption2 <- lm(corruption ~ libdem + polarization + log_wealth, data = modelData) 

summary(corruption2)
```

# Categorical Variables

## Judicial Review and Democracy

<br>

**Judicial Review:**

- Do high courts (Supreme Court, Constitutional Court, etc) have the power to rule on whether laws or policies are constitutional/legal? (Yes or No)
- Dimension of Judicial Independence

## Judicial Review and Democracy

```{r}
#| label: judicial_review

ggplot(modelData, aes(x = log_wealth, y = libdem, color=factor(judicial_review))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "GDP per capita", y = "Liberal Democracy Index") +
  theme_bw() +
    scale_color_manual(name = "Judicial Review", values = c("steelblue3", "coral"), labels = c("No", "Yes")) 
```

## Judicial Review and Democracy

<br>

```{r}
#| label: judicial_review_model
#| echo: true

jr_dem_model <- lm(libdem ~ factor(judicial_review), data = modelData)

summary(jr_dem_model)
```

## Judicial Review and Democracy {.smaller}

<br>

$$\widehat{Democracy_{i}} = 0.17 + 0.28*JudicialReview(yes)$$

- **Slope:** countries with judicial review are expected, on average, to be 0.28 units more democratic on the liberal democracy index
  - Compares baseline level (Judicial Review = 0) to the other level (Judicial Review = 1)
- **Intercept:** average democracy score of countries without judicial review
- Average democracy score of countries with judicial review is 0.17 + 0.28 = 0.45

## Dummy Variables

<br>

- When the categorical explanatory variable has many levels, they're encoded to **dummy variables** 
- We always leave one category out of the model, as the omitted reference category
- Each coefficient describes is the expected difference between level of the factor **and the baseline level**
- Everything is relative to the omitted reference category

## Democracy and World Region

<br>

- Does region predict levels of democracy?
- Since Eastern Europe is the first category, default in R is to use that as the omitted category in models.

```{r}
#| label: region_model
#| echo: true

levels(modelData$region)
```

## Democracy and World Region {.smaller}

How should we interpret intercept?  How about the coefficient on Latin America?

<br>

```{r}
#| label: region_model2
#| echo: true

regional_model1 <- lm(libdem ~ region, data = modelData)

summary(regional_model1)
```

## Democracy and World Region {.smaller}

<br>

What if you want a different baseline category? How do we interpret now?

<br>


```{r}
#| label: region_model3
#| echo: true

# make SS Africa the reference category
modelData <- modelData |>
  mutate(newReg = relevel(region, ref = "SSAfrica"))

regional_model2 <- lm(libdem ~ newReg, data = modelData)

summary(regional_model2)
```

## Your Turn

<br>

**Which types of regime have more corruption?**

<br>
 
V-Dem also includes a categorial regime variable: Closed autocracy (0), Electoral Autocracy (1), Electoral Democracy (2), Liberal Democracy (3)

## Your Turn

<br>

**Which types of regime have more corruption?**

<br>

 
First, let's make this an easier factor variable to work with.

```{r}
#| label: regime_factor
#| echo: true

# Make nicer regime factor variable
modelData <- modelData |> 
  mutate(regime = factor(regime,
                         labels = c("Closed Autocracy",
                                   "Electoral Autocracy",
                                  "Electoral Democracy",
                                  "Liberal Democracy")))
levels(modelData$regime)
```

## Your Turn {.smaller}

<br>

**Which types of regime have more corruption?**

<br>

- Filter data to include only the year 2019 (or run the code to use modelData)
- Make a plot to visualize the relationship between regime type and level of corruption.
  - Which kind of plot is best in this situation?
- Fit a linear model
- Interpret the intercept and coefficients


```{r}
#| label: time2
#| echo: false

library(countdown)

countdown(minutes = 10, 
          id = "timer2", 
          bottom = "10%", 
          right = "10%")
```

## Visualization

```{r}
#| label: corruption_plot

modelData |> 
  group_by(regime) |> 
  summarize(corr = mean(corruption)) |> 
  ggplot(aes(y=corr, x = regime)) +
    geom_col(fill="steelblue") +
  theme_bw()
```

## Model

<br>

```{r}
#| label: corruption_model

corruption <- lm(corruption ~ regime, data = modelData)

summary(corruption)
```

## Create Your Own Model {.smaller}

<br>

- What is a theory that you would like to test with V-Dem data?
- What is the dependent variable?
- What are the independent variables?
- Map out steps to wrangle the data and fit a regression model
- What do you expect to find?
- Now go ahead and wrangle the data
- Fit the model
- Interpret the coefficients and their significance
- Did the results match your expectations?

# Outliers and Model Diagnostics

## What Are Outliers?

<br>

- Observations that don't fit the general pattern of the data
- Not always errors — often **substantively interesting cases**
- In our GDP-democracy example: oil-rich authoritarian states

## Three Types {.smaller}

<br>

- **Leverage points**: extreme values on the x-axis
  - Unusual predictor values — *potential* to distort the model
- **Influential points**: significantly change the regression line when removed
  - Combine extreme predictors *and* large residuals
- **Residual outliers**: points far from the regression line
  - Poorly fit by the model regardless of their x-value

## Regression Diagnostics

<br>

`broom::augment()` adds diagnostic columns to the data:

| Column | What it measures |
|--------|-----------------|
| `.hat` | **Leverage** — how extreme the predictor values are |
| `.std.resid` | **Standardized residuals** — how far from the line |
| `.cooksd` | **Cook's distance** — overall influence on the model |

## Flagging Influential Points {.smaller}

<br>

```{r}
#| label: diagnostics
#| echo: true

diag_data <- vdem |>
  filter(year == 2006) |>
  select(country = country_name, lib_dem = v2x_libdem,
         wealth = e_gdppc, polarization = v2cacamps) |>
  mutate(log_wealth = log(wealth)) |>
  filter(!is.na(lib_dem), !is.na(wealth))

dem_model <- lm(lib_dem ~ log_wealth + polarization, data = diag_data)

model_diagnostics <- augment(dem_model, data = diag_data) |>
  mutate(
    high_leverage  = .hat > 2 * mean(.hat),
    high_residual  = abs(.std.resid) > 2,
    high_influence = .cooksd > 4 / nrow(diag_data)
  )

model_diagnostics |>
  filter(high_leverage | high_residual | high_influence) |>
  select(country, wealth, lib_dem, high_leverage, high_residual, high_influence)
```

## Outlier Strategies

<br>

::: {.incremental}
1. **Leave them in** — outliers may be substantively important
2. **Remove them** — check how much they change your results
3. **Transform the data** — e.g., log transformation compresses extreme values
:::

## Compare Models With and Without Outliers {.smaller}

<br>

```{r}
#| label: compare-outliers
#| echo: true

model_full <- lm(lib_dem ~ log_wealth + polarization, data = diag_data)

model_clean <- model_diagnostics |>
  filter(!high_leverage & !high_residual & !high_influence) |>
  lm(lib_dem ~ log_wealth + polarization, data = _)

tibble(
  model        = c("Full model", "Outliers removed"),
  log_wealth   = c(coef(model_full)[2], coef(model_clean)[2]),
  polarization = c(coef(model_full)[3], coef(model_clean)[3]),
  r_squared    = c(glance(model_full)$r.squared, glance(model_clean)$r.squared)
)
```

## Your Turn! {.smaller}

<br>

- Using `model2` or `model3` from earlier in the session, run `augment()` to identify influential points
- Which countries are flagged? Do they make substantive sense?
- Remove the influential observations and re-fit the model
- How much do the coefficients change? What does that tell you about the robustness of your results?

```{r}
#| label: timer3

library(countdown)

countdown(minutes = 10, id = "timer3", right = "10%", bottom = "5%")
```