---
title: "Lab 1"
subtitle: "Your First Data Visualizations"
author: "YOUR NAME HERE"
format: html
---

::: {.callout-note}
## Getting the Lab Files

Open the Lab 1 project in the [course space on Posit Cloud](https://posit.cloud/spaces/787651) to get started.
:::

::: {.callout-tip title="How to complete this lab"}
Write your code in each empty code chunk and answer the questions below. When you are done, render to HTML to preview your work. For your final submission, change `format: html` to `format: pdf` in the YAML header above and render to PDF.
:::

## Overview

In this lab, you will practice adapting code from the lectures to work with a new dataset. You will:

1.  Load and explore new datasets
2.  Create a bar chart by adapting code from Lesson 1.2
3.  Create a line chart
4.  Write brief interpretations of your visualizations
5.  Render your document to PDF and submit

You are encouraged to have the lecture materials open while completing this lab.

## Getting Started

Load the tidyverse package (make sure that you have it installed).

```{r}
#| label: setup
#| message: false

library(tidyverse)
```

## The Data

Today we'll work with data from the Gapminder project, which tracks development indicators across countries. We have two files:

-   `gapminder_summary.csv` - Average values by continent
-   `gapminder_full.csv` - Data for 142 countries from 1952 to 2007

## Part 1: Bar Chart (50 points)

### Step 1: Load and Explore the Data (10 pts)

First, load the summary dataset and use `glimpse()` to see what variables are available.

```{r}
#| label: load-summary

# Load the data (this is done for you)
gapminder_summary <- read_csv("gapminder_summary.csv")
```

```{r}
#| label: explore-summary

# Use glimpse() to explore the data

```

**Question:** What variables are in this dataset? How many columns are there? How many rows?

YOUR ANSWER HERE

### Step 2: Create a Bar Chart (20 pts)

Using the bar chart code from **Lesson 1.2** as a template, create a bar chart using one of the variables in the dataframe. Your options are `avg_life_exp`, `avg_gdp_cap`, `total_pop`, or `n_countries`.

Think about what needs to change:

- What is the name of your data frame?
- What variable should go on the x-axis?
- What variable should go on the y-axis?
- What should your title and axis labels say?

```{r}
#| label: bar-chart

# Write your bar chart code here

```

### Step 3: Interpret Your Chart (10 pts)

Write 2-3 sentences describing what you see. Which continent has the highest value of the variable you chose to visualize? Which has the lowest?

YOUR INTERPRETATION HERE

## Part 2: Line Chart (50 pts)

### Step 1: Load and Explore the Data (10 pts)

Load the full Gapminder dataset and use `glimpse()` to explore it.

```{r}
#| label: load-gapminder

gapminder <- read_csv("gapminder_full.csv")
```

```{r}
#| label: explore-gapminder

# Use glimpse() to explore the data

```

**Question:** What variables are in this dataset? What years are covered? What does each row represent?

YOUR ANSWER HERE

### Step 2: Filter the Data (10 pts)

Choose three countries you're interested in and filter the data to include only those countries. Replace the `___` placeholders with real country names — spelling must match exactly what you saw in `glimpse()`.

```{r}
#| label: filter-countries
#| eval: false

gapminder_ctrs <- gapminder |>
  filter(country %in% c("___", "___", "___"))
```

### Step 3: Create a Line Chart (20 pts)

Using the line chart code from **Lesson 1.2** as a template, create a line chart showing how life expectancy (`life_exp`) has changed over time for your three countries. Use `year` on the x-axis and `life_exp` on the y-axis.

```{r}
#| label: line-chart

# Write your line chart code here

```

### Step 4: Interpret Your Chart (10 pts)

Write 2-3 sentences describing what you see. How has life expectancy changed over time in your chosen countries? Are there differences across them?

YOUR INTERPRETATION HERE

::: {.callout-important}
## Render as PDF and Submit Your Work (20 pts)

1.  Replace "YOUR NAME HERE" at the top with your actual name
2.  Make sure all your code chunks run without errors
3.  Change `format: html` to `format: pdf` in the YAML header
4.  Click "Render" to create your PDF
5.  Submit the PDF to Blackboard
:::

## Hints

*Only look at these if you're stuck!*

**Hint 1 - Bar chart structure:**

```
ggplot(data_name, aes(x = ___, y = ___)) +
  geom_col() +
  labs(title = "___", x = "___", y = "___")
```

**Hint 2 - Line chart structure:**

```
ggplot(data_name, aes(x = year, y = life_exp, color = country)) +
  geom_line(linewidth = 1)
```

**Hint 3 - Variable names matter:** Make sure the variable names in your code match *exactly* what you saw in `glimpse()`. R is case-sensitive!
