---
title: "Lab 5"
subtitle: "Linear Regression"
author: "YOUR NAME HERE"
format: html
params:
  completed: false
---

::: {.callout-note}
## Getting the Lab Files

Open the Lab 5 project in the [course space on Posit Cloud](https://posit.cloud/spaces/787651) to get started.
:::

::: {.callout-tip title="How to complete this lab"}
Fill in each `???` with the correct code. Once **all** placeholders are filled in, change `completed: false` to `completed: true` in the YAML header above and render to HTML. For your final submission, change `format: html` to `format: pdf`.
:::

## Overview

In this lab, you will practice fitting and interpreting linear regression models using V-Dem democracy data. You will:

1. Wrangle V-Dem data and visualize a bivariate relationship with a regression line
2. Fit a simple linear regression model and interpret the output
3. Extend to multiple regression with two continuous predictors and a categorical predictor
4. Identify and assess the influence of outliers on your model

You are encouraged to have Modules 5.1 and 5.2 open while completing this lab.

## Getting Started

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(vdemdata)
library(broom)
run <- isTRUE(params$completed)
```

::: {.callout-note}
## Installing vdemdata locally

If you are working on your own computer and don't have `vdemdata` installed, you can install it from GitHub. First install the `pak` package, then use it to install `vdemdata`:

```r
install.packages("pak")
pak::pak("vdeminstitute/vdemdata")
```
:::

## The Data

You will work with cross-national data from the V-Dem project for the year 2019. Parts 1 and 2 focus on a simple research question: **Are more democratic countries less corrupt?**

The key variables are:

- `lib_dem` — Liberal Democracy Index (0–1, higher = more democratic)
- `corruption` — Political Corruption Index (0–1, higher = more corrupt)

## Part 1: Wrangle and Visualize (30 points)

### Step 1: Wrangle the data (10 pts)

Fill in the `???` to filter V-Dem to the year 2019 and select the relevant variables.

```{r}
#| label: wrangle
#| eval: !expr run

model_data <- vdem |>
  filter(year == ???) |>
  select(
    country = country_name,
    lib_dem = v2x_libdem,
    corruption = v2x_corr
  )

glimpse(model_data)
```

**Question:** How many rows does `model_data` have? What does each row represent?

YOUR ANSWER HERE

### Step 2: Visualize the relationship (20 pts)

Create a scatter plot with `lib_dem` on the x-axis and `corruption` on the y-axis. Add a linear regression line using `geom_smooth()`. Include a title and axis labels, and use `theme_bw()`.

```{r}
#| label: scatterplot

# Write your plot code here

```

**Question:** Describe the relationship. Is it positive or negative? Does it look roughly linear?

YOUR ANSWER HERE

## Part 2: Simple Linear Regression (50 points)

### Step 1: Fit the model (10 pts)

Use `lm()` to fit a simple linear regression predicting `corruption` from `lib_dem`. Store the result in an object called `model1` and view the output with `summary()`.

```{r}
#| label: fit-model
#| eval: !expr run

model1 <- lm(??? ~ ???, data = model_data)

summary(model1)
```

### Step 2: Write out the equation (10 pts)

Using the `Estimate` column from the output, write out the estimated regression equation below (round coefficients to two decimal places):

$$\widehat{Corruption}_i = a + b \times Democracy_i$$

YOUR EQUATION HERE

### Step 3: Interpret the intercept (10 pts)

**Question:** What is the value of the intercept, and what does it represent in the context of this model?

YOUR ANSWER HERE

### Step 4: Interpret the slope (15 pts)

**Question:** What is the value of the slope, and what does it tell us about the relationship between democracy and corruption? Be specific about direction and magnitude.

YOUR ANSWER HERE

### Step 5: Statistical significance (5 pts)

**Question:** Is the slope statistically significant at the 0.05 level? How do you know from the `summary()` output?

YOUR ANSWER HERE

## Part 3: Multiple Regression (50 points)

Now we extend the analysis. The new research question is: **What predicts a country's level of liberal democracy?**

The key variables are:

- `lib_dem` — Liberal Democracy Index (0–1, higher = more democratic)
- `wealth` — GDP per capita (in thousands of USD)
- `log_wealth` — log-transformed GDP per capita
- `polarization` — Political polarization index (higher = more polarized)
- `region` — World region (6 categories)

### Step 1: Wrangle the data (10 pts)

Fill in the `???` to filter V-Dem to the year 2019 and select the relevant variables.

```{r}
#| label: wrangle-multi
#| eval: !expr run

model_data2 <- vdem |>
  filter(year == ???) |>
  select(
    country = country_name,
    lib_dem = v2x_libdem,
    wealth  = e_gdppc,
    polarization = v2cacamps,
    region = e_regionpol_6C
  ) |>
  mutate(
    log_wealth = log(wealth),
    region = factor(region,
      labels = c("Eastern Europe", "Latin America", "MENA",
                 "Sub-Saharan Africa", "Western Europe & N. America",
                 "Asia and Pacific"))
  ) |>
  filter(!is.na(lib_dem), !is.na(wealth))

glimpse(model_data2)
```

**Question:** How many countries are in the dataset after filtering out missing values?

YOUR ANSWER HERE

### Step 2: Fit a model with two predictors (20 pts)

Fit a multiple regression model predicting `lib_dem` from `log_wealth` and `polarization`. Store the result as `model2` and view the output with `summary()`.

```{r}
#| label: model2
#| eval: !expr run

model2 <- lm(??? ~ ??? + ???, data = model_data2)

summary(model2)
```

**Question:** Write out the estimated regression equation using the coefficients from the output (round to two decimal places).

$$\widehat{LibDem}_i = a + b_1 \times LogWealth + b_2 \times Polarization$$

YOUR EQUATION HERE

**Question:** Interpret the coefficient on `log_wealth`. What does "holding polarization constant" mean in plain language?

YOUR ANSWER HERE

**Question:** Interpret the coefficient on `polarization`. Is the direction of the relationship what you would expect? Why or why not?

YOUR ANSWER HERE

### Step 3: Add a categorical predictor (20 pts)

Now add `region` as a third predictor. Store this as `model3` and view the summary.

```{r}
#| label: model3
#| eval: !expr run

model3 <- lm(lib_dem ~ log_wealth + polarization + region, data = model_data2)

summary(model3)
```

**Question:** What is the baseline (reference) region in this model? How do you know?

YOUR ANSWER HERE

**Question:** Pick one region coefficient and interpret it in a sentence. What does it tell us about democracy in that region compared to the baseline?

YOUR ANSWER HERE

## Part 4: Outliers and Influential Observations (30 points)

### Step 1: Identify influential points (15 pts)

Use `broom::augment()` to compute regression diagnostics for `model2`. Then flag observations that are high-leverage, high-residual, or highly influential.

```{r}
#| label: diagnostics
#| eval: !expr run

model_diagnostics <- augment(model2, data = model_data2) |>
  mutate(
    high_leverage  = .hat > 2 * mean(.hat),
    high_residual  = abs(.std.resid) > 2,
    high_influence = .cooksd > 4 / nrow(model_data2)
  )

model_diagnostics |>
  filter(high_leverage | high_residual | high_influence) |>
  select(country, wealth, lib_dem, high_leverage, high_residual, high_influence)
```

**Question:** How many countries are flagged as influential? Name two that stand out and explain why they might be outliers in this model.

YOUR ANSWER HERE

### Step 2: Compare models with and without influential points (15 pts)

Remove the influential observations and re-fit the model. Compare the coefficients and R-squared between the two versions.

```{r}
#| label: compare-models
#| eval: !expr run

model_full <- lm(lib_dem ~ log_wealth + polarization, data = model_data2)

model_clean <- model_diagnostics |>
  filter(!high_leverage & !high_residual & !high_influence) |>
  lm(lib_dem ~ log_wealth + polarization, data = _)

tibble(
  model        = c("Full model", "Outliers removed"),
  intercept    = c(coef(model_full)[1],    coef(model_clean)[1]),
  log_wealth   = c(coef(model_full)[2],    coef(model_clean)[2]),
  polarization = c(coef(model_full)[3],    coef(model_clean)[3]),
  r_squared    = c(glance(model_full)$r.squared, glance(model_clean)$r.squared)
)
```

**Question:** How much do the coefficients change after removing influential points? Does this suggest the outliers were having a large or small effect on the results? What does this tell you about the robustness of the findings?

YOUR ANSWER HERE

::: {.callout-important}
## Render as PDF and Submit Your Work

1. Replace "YOUR NAME HERE" at the top with your actual name
2. Make sure all code chunks run without errors
3. Click "Render" to create your PDF
4. Submit the PDF to Blackboard
:::

---

## Bonus A: Build Your Own Model (up to 10 points)

Choose a different outcome variable and predictor from V-Dem that you find theoretically interesting. Wrangle the data, create a scatter plot with a regression line, fit the model, and interpret the slope and intercept.

```{r}
#| label: your-model

# Write your code here

```

**Question:** What variables did you choose and why? Interpret the slope and intercept, and describe whether the results matched your expectations.

YOUR ANSWER HERE

---

## Bonus B: Checking Model Assumptions (up to 10 points)

*This section uses the `performance` package to run automated diagnostics.*

```{r}
#| label: check-model
#| message: false
#| eval: !expr run

library(performance)

check_model(model3)
```

**Question:** Look at the **Linearity** panel. Does the smoother line stay close to zero across fitted values, or does it show a curve? What would a curve indicate?

YOUR ANSWER HERE

**Question:** Look at the **Homogeneity of Variance** panel. Does the spread of residuals look roughly constant, or does it fan out? What would a funnel shape indicate?

YOUR ANSWER HERE

**Question:** Look at the **Normality of Residuals** panel. Do the points follow the diagonal line closely? Are there any systematic departures in the tails?

YOUR ANSWER HERE

---

## Hints

*Only look at these if you're stuck!*

**Hint 1 — Filtering to one year:**

```
filter(year == 2019)
```

**Hint 2 — Scatter plot with regression line:**

```r
ggplot(model_data, aes(x = lib_dem, y = corruption)) +
  geom_point() +
  geom_smooth(method = "lm", color = "#E48957", se = FALSE) +
  labs(
    title = "Democracy and Corruption, 2019",
    x = "Liberal Democracy Index",
    y = "Corruption Index"
  ) +
  theme_bw()
```

**Hint 3 — Fitting a simple linear model:**

```r
model1 <- lm(corruption ~ lib_dem, data = model_data)
summary(model1)
```

**Hint 4 — Reading the summary output:**

The `Estimate` column gives the intercept and slope. The `Pr(>|t|)` column gives the p-value — values below 0.05 are statistically significant at the 5% level.

**Hint 5 — Fitting a multiple regression model:**

```r
model2 <- lm(lib_dem ~ log_wealth + polarization, data = model_data2)
summary(model2)
```

**Hint 6 — Including a factor variable as a predictor:**

```r
model3 <- lm(lib_dem ~ log_wealth + polarization + region, data = model_data2)
```

R automatically creates dummy variables for each level of `region`, leaving the first level as the reference category.

**Hint 7 — Running regression diagnostics:**

```r
model_diagnostics <- augment(model2, data = model_data2) |>
  mutate(
    high_leverage  = .hat > 2 * mean(.hat),
    high_residual  = abs(.std.resid) > 2,
    high_influence = .cooksd > 4 / nrow(model_data2)
  )
```

**Hint 8 — Removing outliers and re-fitting:**

```r
model_clean <- model_diagnostics |>
  filter(!high_leverage & !high_residual & !high_influence) |>
  lm(lib_dem ~ log_wealth + polarization, data = _)
```

**Hint 9 — Comparing coefficients with `glance()` and `coef()`:**

`coef(model)` extracts the coefficients as a named vector. `glance(model)$r.squared` extracts R-squared from the `broom` package.
