---
title: "Lab 2"
subtitle: "Wrangling and Exploring Democracy Data"
author: "YOUR NAME HERE"
format: html
params:
  completed: false
---

::: {.callout-note}
## Getting the Lab Files

Open the Lab 2 project in the [course space on Posit Cloud](https://posit.cloud/spaces/787651) to get started.
:::

## Overview

In this lab, you will practice the main skills from Modules 2.1 and 2.2. You will:

1. Load and wrangle V-Dem data
2. Summarize democracy indicators by region
3. Create a column chart from summarized data
4. Explore categorical regime data with `geom_bar()`
5. Compare regime proportions across regions
6. Write brief interpretations of your results

Fill in each `???` with the correct code. You are encouraged to have the Week 2 lecture materials and the V-Dem codebook open while completing the lab.

## Getting Started

Load the required packages.

```{r}
#| label: setup
#| message: false

library(vdemdata)
library(tidyverse)
run <- isTRUE(params$completed)
```

::: {.callout-note}
## Installing `vdemdata` locally

If you are working on your own computer and do not have `vdemdata` installed, install `pak` and then install `vdemdata` from GitHub:

```r
install.packages("pak")
pak::pkg_install("vdeminstitute/vdemdata")
```
:::

## Part 1: Wrangle Democracy Data (30 points)

In this part, you will build a clean dataset from V-Dem using `filter()`, `select()`, and `mutate()`.

### Step 1: Create a Wrangled Dataset (20 pts)

Filter the data to one year, select the variables you need, and recode the region variable into readable labels.

```{r}
#| label: wrangle-democracy
#| eval: !expr run

democracy <- vdem |>
  ???(year == ???) |>
  ???(
    country = ???,
    year,
    polyarchy = ???,
    libdem = ???,
    gdp_pc = ???,
    region = ???
  ) |>
  ???(
    region = case_match(region,
      1 ~ "Eastern Europe",
      2 ~ "Latin America",
      3 ~ "Middle East",
      4 ~ "Africa",
      5 ~ "The West",
      6 ~ "Asia")
  )
```

### Step 2: Examine the Data (5 pts)

Run `glimpse()` on your new data frame.

```{r}
#| label: democracy-glimpse
#| eval: !expr run

glimpse(???)
```

### Step 3: Brief Response (5 pts)

Write 2-3 sentences answering these questions:

- What year did you choose?
- How many rows and columns does your wrangled dataset have?
- Why is it useful to recode `region` before plotting?

YOUR RESPONSE HERE

## Part 2: Summarize and Visualize by Region (35 points)

In this part, you will practice `group_by()`, `summarize()`, `arrange()`, and `geom_col()`.

### Step 1: Summarize the Data (15 pts)

Create a regional summary dataset. Use `mean()` for `polyarchy` and `gdp_pc`, and sort from highest to lowest `polyarchy`.

```{r}
#| label: summarize-region
#| eval: !expr run

dem_summary <- democracy |>
  ???(region) |>
  ???(
    polyarchy = ???(polyarchy, na.rm = TRUE),
    gdp_pc = ???(gdp_pc, na.rm = TRUE)
  ) |>
  ???(desc(polyarchy))
```

### Step 2: Print the Summary (5 pts)

Print your summarized data frame.

```{r}
#| label: print-summary
#| eval: !expr run

???
```

### Step 3: Create a Column Chart (10 pts)

Create a column chart of average polyarchy by region. Reorder the bars from highest to lowest.

```{r}
#| label: col-chart

# Write your code here

```

### Step 4: Brief Interpretation (5 pts)

Write 2-3 sentences describing what you see.

- Which region has the highest average polyarchy score?
- Which region has the lowest?
- Does the ranking make sense to you?

YOUR RESPONSE HERE

## Part 3: Explore Categorical Regime Data (35 points)

In this part, you will work with V-Dem's `v2x_regime` variable and use `count()` and `geom_bar()`.

### Step 1: Build a Regime Dataset (15 pts)

Filter the data to 2022, select `country`, `regime`, and `region`, and recode both `region` and `regime`.

```{r}
#| label: regime-setup
#| eval: !expr run

vdem2022_regime <- vdem |>
  ???(year == ???) |>
  ???(
    country = ???,
    regime = ???,
    region = ???
  ) |>
  ???(
    region = case_match(region,
      1 ~ "Eastern Europe",
      2 ~ "Latin America",
      3 ~ "Middle East",
      4 ~ "Africa",
      5 ~ "The West",
      6 ~ "Asia"),
    regime = case_match(regime,
      0 ~ "Closed Autocracy",
      1 ~ "Electoral Autocracy",
      2 ~ "Electoral Democracy",
      3 ~ "Liberal Democracy")
  )
```

### Step 2: Count Regime Types (5 pts)

Create a frequency table of regime types.

```{r}
#| label: regime-count
#| eval: !expr run

vdem2022_regime |>
  ???(regime)
```

### Step 3: Create a Bar Chart of Regime Types (5 pts)

Use `geom_bar()` to plot the distribution of regime types.

```{r}
#| label: regime-bar

# Write your code here

```

### Step 4: Compare Regions with Proportions (5 pts)

Now create a second bar chart comparing regime types across regions using `position = "fill"`.

```{r}
#| label: regime-proportion-bar

# Write your code here

```

### Step 5: Brief Interpretation (5 pts)

Write 2-3 sentences describing what you see.

- Which regime type is most common overall?
- Which region appears to have the largest share of liberal democracies?
- Why is `position = "fill"` useful here?

YOUR RESPONSE HERE

::: {.callout-important}
## Submission

1. Replace "YOUR NAME HERE" at the top with your actual name
2. Make sure all code chunks run without errors
3. Change `completed: false` to `completed: true`
4. Render to HTML to check your work
5. Change `format: html` to `format: pdf`
6. Submit the PDF to Blackboard
:::

## Hints

*Only look at these if you are stuck.*

**Hint 1 - Wrangling structure**

```r
new_data <- vdem |>
  filter(year == 2015) |>
  select(
    country = country_name,
    year,
    polyarchy = v2x_polyarchy
  ) |>
  mutate(...)
```

**Hint 2 - Summarizing structure**

```r
summary_data <- democracy |>
  group_by(region) |>
  summarize(
    polyarchy = mean(polyarchy, na.rm = TRUE)
  ) |>
  arrange(desc(polyarchy))
```

**Hint 3 - `geom_col()`**

```r
ggplot(dem_summary, aes(x = reorder(region, -polyarchy), y = polyarchy)) +
  geom_col()
```

**Hint 4 - `geom_bar()`**

```r
ggplot(vdem2022_regime, aes(x = regime)) +
  geom_bar()
```

**Hint 5 - Proportional bar chart**

```r
ggplot(vdem2022_regime, aes(x = region, fill = regime)) +
  geom_bar(position = "fill")
```
